A Speculative Approach to Parallelization in Particle Swarm Optimization

### 2011-02-08 honors_thesis/si_paper/reviewer_response/response.tex

\section*{Associate Editor Comments}

\subsection*{AE.1}

Reviewer comments: The quality of the article will be improved if you include a
review of existing parallelization methods for PSO, to produce a kind of
taxonomy. Then, also include comparisons with these approaches. You currently
compare only with a naive approach and adaptations of your approach.

Response: We added a Related Work section (now section 3), describing previous
approaches at parallelizing PSO.  We also added (NOT DONE) experiments
comparing against another common parallelization approach, that of using
subswarms.

@@

NOT DONE

\subsection*{AE.3}

Reviewer comments: It will also be good if you do make use of problems where
parallelization has benefit. For the functions that you use, there is not real
not for parallelization. Linked to this, when will parallelization be
beneficial?

Response: We added a paragraph to what is now Section 7 addressing the issue of
benchmark functions (``We also do not report\ldots'').  We also ran a
comparison on a longer-running function and included it in our results (NOT
DONE).

@@

NOT DONE

\subsection*{AE.4}

Reviewer comments: Please re-organize section 3: You start with a very long
theoretical discussion, and only then get to section 3.1.

Response: Good point.  We realized that we tried to cram too much into a single
section, so we split the two main parts of it into two separate sections.
There is now a section devoted entirely to the mathematics of the method (now
Section 4) and a separate section describing its implementation (Section 5).
We also put subsections in the new Section 4 to guide the reader a little
better through the section.

@@

Response: This is a fair criticism, as it is true that experimentation with a
problem is required before one can know how best to solve it.  We added two
paragraphs to the end of section 2 to further make this point, making more
prominent reference to the No Free Lunch theorems.

\subsection*{1.2}

Reviewer comments: The organization and presentation of the ideas can be
greatly improved. The authors mix results and conclusions with the description
of the approach. This doesn't help the goal of the authors. A reader, cannot
appreciate the significance of some result or conclusion if he/she hasn't
understood what the authors are talking about. I recommend the authors to
restructure the whole paper but in particular Sections 3 and 4.

Response: We substantially reorganized Section 3, as noted above, and we moved
the discussion of branch statistics (the old Section 4.3) to the beginning of
the results section (NOT DONE).

@@

NOT DONE

\subsection*{1.3}

Reviewer comments: The notation used throughout the paper is confusing. In the
PSO literature, it is common to use x as a particles current position and some
other letters to denote a particle's personal best position and a particle's
neighborhood's best position. I know that the authors are using superscripts to
distinguish between these concepts but they don't really help.

Response: We had a long discussion about notation before writing our paper, and
we thought that because the personal best and neighborhood bests are positions,
using an x to denote it would be more intuitive.  However, we see your point
that it can be confusing for other reasons.  We changed the x to a b when
referring to best positions, while still using the N and P superscripts to
differentiate between which best is being referred to.

@@

\subsection*{1.4}

Reviewer comments: Using ``rounds of evaluations'' is confusing, and results
should be reported in terms of wall clock time.

Response: Rounds of evaluations correspond exactly to wall clock time, as we
describe in the second paragraph of what was Section 5 (now Section 7).
However, we realize that the term is not a commonly used one, so we changed
``rounds of evaluations'' to ``time steps'' in hopes of better conveying our
meaning.  We also added some text in the description mentioned above that
hopefully further clarifies this point.

@@

\subsection*{1.5}

Reviewer comments: Since the authors are using Herrera et al.'s proposed
benchmarks and since they are using a large computing cluster, I don't see why
they can't use the same experimental protocol proposed by Herrera et al., which
is focused on large scale optimization. That protocol would better justify a
parallel algorithm because a 1000-dimensional problem is intuitively more
difficult than the 20-dimensional problems the authors are currently using. As
a side product, the authors can actually compare the results of the pruned
versions (which in my eyes, are new PSO variants) with those of other
state-of-the-art algorithms.

Response: Running a few more sets of experiments at higher dimensionalities
should address this adequately, though we can't really do all of their
protocol.

@@

NOT DONE

\subsection*{1.6}

Reviewer comments: In the abstract, it is not clear the meaning of
``speculative''. A reader without knowledge about parallel computing may not
understand what the authors mean.

Response: We had a paragraph explaining this in the introduction, though the
analogy to processors lacked an explicit connection to the word
``speculative.''  We added a sentence at the end of that paragraph to address
this issue.

@@

Reviewer comments: The authors can cite Poli's paper about applications of PSO
algorithms in the first paragraph of the paper.

Response: We added the citation after the first sentence of the paper.

\subsection*{1.8}

Reviewer comments: What serves as the motivation for the authors's work is
insufficiently discussed.  The authors say that adding more particles leads to
``diminishing returns''. This may be true when dealing with constant size
populations.  However, a recent trend is to vary the population size over time.
In fact, it has been shown that adding particles \emph{over time} may actually
solve the problem the authors refer to. Some references are ``Incremental
Social Learning in Particle Swarms'' by Montes de Oca and others published in
IEEE SMCB and ``Efficient Population Utilization Strategy for Particle Swarm
Optimizer'' by Hsieh and others in the same journal.

@@

Response: See how papers are relevant (Andrew).  But also, assuming a fixed
computational budget, if at any point you have a smaller swarm than your budget
allows, you should do speculative evaluation.  That's our point.

Put those papers in related work, on another way to decide what computation to
perform---some people increase swarm size over time.  Relate that to our point.

@@

NOT DONE

\subsection*{1.9}

Reviewer comments: In the first paragraph of Section 3, the authors say that
adding particles may not help past a certain swarm size. This is true if the
computation time allocated to the optimization process remains fixed. However,
in some problems, it is the solution quality what matters. In that case, adding
particles \emph{and} computational budget may indeed be the best thing to do.

Response: We added some text to to the introduction (paragraph beginning with
``In this paper we consider\ldots'') to better explain our intent; that is,
given a fixed computational budget, how can processors be used most
efficiently?  However, it is also the case that if there is ever a point in
your computation where you have a bigger computational budget than the
particles themselves demand, speculative evaluation is the best way to utilize
the remaining computational budget.  We added text to the conclusion to explain
this point (or maybe to related work?).

@@

NOT DONE

\subsection*{1.10}

Reviewer comments: In Eq. 3. Why must the current position be worse than the
neighbor's position?

Response: Your question probably arose from the confusing notation.  We have
fixed the notation and added a parenthetical statement directly preceeding the
equation to better explain what the equation is.  The equation is a specific
example of a possible update case, and in the case described the neighborhood
best is not updated; thus by definition the function values of current
positions must not be lower than the value of the previous neighborhood best.

@@

Response: We fixed section 3 by splitting it into two sections and reorganizing
some parts, making the examples that were there more clear.  Hopefully the
section is satisfactory now.

\subsection*{1.12}

Reviewer comments: Isn't pruning against the idea of better using the available
computing resources? If one reduces the number of evaluated speculative
children, then fewer CPUs are going to be used.

Response: Pruning allows for the use of a larger swarm with the same number of
processors.  We realized that this was never explicitly stated in the text, so
we added a few sentences to the description of pruning (now section 6.2) to
make it clear.  The text added to the introduction explaining our purpose of
fully utilizing resources also addresses this issue, though somewhat
indirectly.

@@

\subsection*{1.13}

Reviewer comments: Lines 29-30 of page 14. Stagnation is about the lack of
improvement of the best result. It has nothing to do with what the particles
are doing. I know that the PSO community confuses the two. I think this is a
nice opportunity to clarify the difference between search stagnation and the
state of the swarm.

Response: In fact the best result is improved, frequently, it is just that the
scale of the improvements becomes increasingly smaller.

@@

NOT DONE

\subsection*{1.15}

Reviewer comments: Page 27, line 40. Algorithms are suited for problems not the
other way around.

Response: Duly noted.  We fixed the sentence (it now says ``Our methods work
very well on these functions'').

@@

\subsection*{1.16}

Reviewer comments: Page 31. Line 46. Molia $\rightarrow$ Molina. Is it a
technical report?  Which number?  Isn't it published with the special issue?

Response:  We fixed the typo, and we looked for quite some time trying to find
a better way to cite that paper.  The special issue hasn't been officially
published yet, from what I can tell, and the editorial that is online (under
``Online First'' at springerlink.com) only has a link to the test suite
document.  I suppose that we could cite the editorial instead of the test suite
document, if you think that would be preferable.

@@

Response: ...

NOT DONE

\subsection*{2.2}

Reviewer comments: In the first sections of the paper it is argued that the
proposed ``...method can easily be extended to arbitrary topologies''. However,
in later sections it is apparent that this may be possible if the underlying
neighborhood topology is not ``dense'' enough to cause heavy communication
traffic to the processors.  Can you be more precise to this and quantify it?
For example, given a swarm size and a common network communication is there any
identifiable limit to the neighborhood topology such that the algorithm retains
an acceptable performance?

@@

NOT DONE

\subsection*{2.3}

Reviewer comments: My personal experience with the PSO versions used in the
paper suggests that, in large dimensions, it is preferrable to use larger
swarms for less iterations rather than small swarms for more iterations. This
is in contrast to the argued behavior of PSO for the Sphere function (Section
1).  Did you observe the same behavior in other functions?  Could you provide
some evidence on this because, due to the swarm's diversity loss, it seems
somehow counterintuitive.

Response: The citation given in the paper (McNabb et al. CEC 09) gives an
interesting discussion on these points.  The functions explored are limited,
but the paper shows that both Sphere and Griewank exhibit this behavior, while
Rastrigin does not (the relevant figures in that paper are 2, 4 and 6).  As
those results are previously published and cited in our paper, we felt it
unnecessary to reduplicate the figures, though we make mention of the
conclusions.  However, our own figures also answer your question.  The fact
that our methods outperform the na\"ive parallelization is in itself evidence
for this claim.  Recall that when our figures say SEPSO, that is equivalent to
running PSO with a swarm size one eighth the size, doing twice as many
iterations.  Figures 9 and 10 illustrate the point that for some swarm sizes
you are correct and it is better to have a larger swarm size.  However, once
enough processors are available to have a large enough swarm even with SEPSO,
it is better by far to do speculative evaluation.

Increasing the swarm size increases the diversity, as you say (the McNabb et
al. paper makes the same point).  But at some point there is enough diversity,
and when that point is reached extra processors should be used for speculative
evaluation.  We spend almost the entire conclusion trying to make this point.

@@

\section*{Reviewer 3 Comments}

\subsection*{3.1}

Reviewer comments: There are too many redundancy sentences declaring this paper
is not proposing a new algorithm in the introduction.

Response: We found only one part of the introduction that mentioned the paper
not proposing a new algorithm.  That part had two sentences, and we deleted the
second one to soften the point a little.  It was an over-reaction to comments
by previous reviewers.

@@

\subsection*{3.2}

Reviewer comments: The velocity updating formular is different from the
original PSO algorithm.  The weighting factor, X, should only affects the old
velocity, not all the items.

Response: You are referring to inertia weight PSO (the original equations in
Kennedy 1995 did not have any weighting factor).  Constricted PSO is also a
widely used improvement on the original PSO, and we cite two papers as
justification for our use of those formulas.  However, our text wasn't clear
that we were using the constricted PSO equations, so we clarified the text.

@@

\subsection*{3.3}

Reviewer comments: Some reference are not completed. For example, the
journal/conference name of the 4th reference is missing. Please check it
carefully.

Response: Thanks for catching that.  We fixed it and went over our other
references carefully.

@@

\subsection*{3.4}

Reviewer comments: Please increasing the reference to extend the integrity of
the background.

Response: The Chu et al. paper that we already cite seems to be an extension of
the Chang et al. paper you mentioned.  However, we included the Chang et al.
paper along with the Chu et al. paper in our review of related work.  We also
included several more related papers in our review of current parallelization
strategies for PSO.

### 2011-02-08 honors_thesis/si_paper/paper.tex
Major revisions to paper in response to feedback

%% PSO Stuff
\DeclareMathOperator*{\argmin}{arg\;min}
\DeclareMathOperator*{\argmax}{arg\;max}
\DeclareMathOperator*{\arginf}{arg\;inf}
\DeclareMathOperator*{\argsup}{arg\;sup}
\providecommand{\pers}{\ensuremath{P}}
\providecommand{\neigh}{\ensuremath{N}}
\providecommand{\leftind}{\ensuremath{L}}
\providecommand{\rightind}{\ensuremath{R}}
\providecommand{\nURand}{\ensuremath{U^\neigh}}
\providecommand{\pURand}{\ensuremath{U^\pers}}
\providecommand{\ppos}{\ensuremath{\Vec{x}}}
\providecommand{\fppos}{\ensuremath{\Vec{fx}}}
\providecommand{\pvel}{\ensuremath{\Vec{v}}}
\providecommand{\nbest}{\ensuremath{\Vec{x}^\neigh}}
\providecommand{\fnbest}{\ensuremath{\Vec{fx}^\neigh}}
\providecommand{\pbest}{\ensuremath{\Vec{x}^\pers}}
\providecommand{\fpbest}{\ensuremath{\Vec{fx}^\pers}}
\providecommand{\constriction}{\ensuremath{\chi}}
\providecommand{\ncoeff}{\ensuremath{\phi^\neigh}}
\providecommand{\pcoeff}{\ensuremath{\phi^\pers}}
\providecommand{\obs}{\ensuremath{\Vec{\xi}}}
\providecommand{\ofunc}{\ensuremath{f}}
\providecommand{\swarm}{\ensuremath{swarm}}

@@

Particle swarm optimization (PSO) has been found to be a highly robust and
effective algorithm for solving many types of optimization
problems~\cite{poli-2008-pso-applications}.  For much of the algorithm's
history, PSO was run serially on a single machine.  However, the world's
computing power is increasingly coming from large clusters of processors.  In
order to efficiently utilize these resources for computationally intensive
problems, PSO needs to run in parallel.

@@

However, it can be seen from the graph that doubling the swarm size produces
only incremental benefit; performance improves, but not by very much.  The
reason for this is that there comes a point of diminishing returns with respect
to adding particles.  In \fig{evals-sphere} we show the value obtained after
50,000 function evaluations (not iterations) as a function of swarm size, again
for the function Sphere.  Increasing the swarm size from 5 to 10 has a
significant effect on the value obtained.  However, increasing the swarm size
from 16 to 30 makes the algorithm less efficient; that is, it reduces the
progress the algorithm makes per evaluation.  Other functions show similar
trends, though often the optimal swarm size is slightly larger.  For this
reason, previous work has recommended the use of a swarm size of 50 for
PSO~\citep{bratton-2007-defining-a-standard-for-pso}.  Thus, in at least some
cases, adding particles indefinitely will not yield an efficient
implementation.

@@

In this paper we consider PSO parallelization strategies for clusters of
hundreds or thousands of processors and functions for which a single evaluation
will take at least tens of seconds, but probably minutes or perhaps hours.  Our
purpose is to explore the question of what to do with a thousand processors
when 50 or 100 particles is the most efficient swarm size, and simply
adding particles results in only incremental improvement.

In order to solve this problem, we apply the concept of speculative
decomposition~\citep{grama-2003-intro-to-parallel-computing} to particle swarm
optimization, using extra processors to perform two iterations of PSO at the
same time.  Speculative decomposition is analogous to speculative execution
(also known as branch prediction), a technique commonly used in processors.
Modern processors, when faced with a branch on which they must wait (e.g., a
memory cache miss), guess which way the branch will go and start executing,
ensuring that any changes can be undone.  If the processor guesses right,
execution is much farther ahead than if it had idly waited on the memory
reference.  If it guesses wrong, execution restarts where it would have been
anyway.

In this paper we show that the results of standard PSO can be reproduced
\emph{exactly}, two iterations at a time, using a speculative approach adapted
from speculative execution. We prove that the standard PSO equation can be
factored such that a set of speculative positions can be found which will
\emph{always} include the position computed in the next iteration.  By
computing the value of the objective function for each of the speculative
positions at the same time the algorithm evaluates the objective function for
the current position, it is possible to know the objective function values for
both the current and the next iteration at the same time.  We demonstrate this
principle by implementation and show that it produces exactly the same results
as standard PSO, but two iterations at a time.  The resulting implementation
runs efficiently on large clusters where the number of processors is much
larger than a typical or reasonable number of particles, producing better
results in less ``wall-clock'' time.

It is important to note here that this is not a variant of PSO.  We are not
proposing a new algorithm, and we do not try to beat the state of the art in
PSO research.  We simply propose a new way to think about the parallelization
of PSO that we show works better in many cases than a na\"ive parallelization.

@@

The balance of this paper is organized as follows. \sec{pso} describes the
particle swarm optimization algorithm.  \sec{sepso} describes mathematically
how speculative evaluation can be done in parallel PSO to perform two
iterations at once, and \sec{implementation} gives information on how to
implement this method.  In \sec{relax}, we discuss various methods of improving
the performance of speculative evaluation in PSO, all of which break the
requirement of strictly reproducing the behavior of the original algorithm.
\sec{setup} describes the experiments we ran, and \sec{results} presents the
results of those experiments.  In \sec{conclusion} and \sec{future} we conclude
and discuss future work.

@@

At each iteration of the algorithm, the position $\ppos_t$ and velocity
$\pvel_t$ of each particle are updated as follows:
\begin{align}
\nonumber
	\pvel_{t+1} &=
		\constriction \bigl[ \pvel_t
			+ \pcoeff\pURand_{t}\otimes(\pbest_{t} - \ppos_{t}) \\
\label{eq:velupdate}
			& \quad \quad \quad \, + \ncoeff\nURand_{t}\otimes(\nbest_{t} - \ppos_{t})
		\bigr] \\
\label{eq:posupdate}
	\ppos_{t+1} &= \ppos_{t} + \pvel_{t+1}
\end{align}
where \( \pURand_{t} \) and \( \nURand_{t} \) are vectors of independent random
numbers drawn from a standard uniform distribution, the \( \otimes \) operator
is an element-wise vector multiplication, $\pbest$ (called personal best) is
the best position the current particle has seen, and $\nbest$ (called
neighborhood best) is the best position the neighbors of the current particle
have seen~\citep{bratton-2007-defining-a-standard-for-pso}.  The parameters \(
\ncoeff \), \( \pcoeff \), and \( \constriction \) are given prescribed values
required to ensure convergence (2.05, 2.05, and .73,
respectively)~\citep{clerc-2002-constricted-pso}.

@@

The ideal topology and swarm size for PSO depend on the objective function.
Researchers have devised various benchmark functions and have found that the
ideal topology for one function may perform very poorly for another function.
The No Free Lunch Theorems for Optimization show that this is true in
general---if an algorithm performs well on average for one class of functions
then it must do poorly on average for other
problems~\cite{wolpert-1997-nfl-for-optimization}.

An attempt to standardize PSO found that although a Complete swarm of 50
particles converged to the global optimum more quickly for many benchmark
functions than the same swarm size with a Ring topology, it was also more
likely to permaturely converge to local optima for other functions.  The study
found no significant improvement for any other swarm size between 20 and 100
and concluded with a recommendation to use a swarm of 50 particles with a Ring
topology as a starting point.  The authors acknowledged that choosing the ideal
topology requires thorough experimentation for the particular
problem~\cite{bratton-2007-defining-a-standard-for-pso}.  Other authors have
explained that a large swarm with a sparse topology propogates information
slowly~\cite{montes-de-oca-2009-frankensteins-pso}.

@@

To do so we must first refactor PSO such that the determination of the value
of the objective function is separate from the rest of the computation.  For
simplicity, this discussion will describe the case where PSO is performing
function minimization using the Ring topology.  In this example, each particle
has two neighbors, the ``right neighbor'' and ``left neighbor,'' whose
positions are represented as $\rightn$ and $\leftn$ respectively.  Though we
will only describe the case of the Ring topology, this method can easily be
extended to arbitrary topologies.

The refactoring hinges on the idea that once the random coefficients
$\pURand_{t}$ and $\nURand_{t}$ are determined, there are only a few possible
new positions, or updates, for $\nbest$ and $\pbest$.  For the Ring topology
there are 7 possible update cases, identified in Table~\ref{tab:evals}.  We
label each case with an identifier referring to the source of the update: a
minus sign ($-$) represents no update, $L$ represents an update to $\nbest$
coming from the left neighbor, $R$ represents an update to $\nbest$ coming from
the right neighbor, and $S$ represents an update to either $\pbest$ or $\nbest$
coming from the particle itself.  As an example, $\casexn$ refers to the case
that the particle finds a new personal best, but neither it nor its neighbors
find a position that updated its neighborhood best.  In the equations that
follow, we refer to an unspecified update case as $\casegen$, and to the set of
cases collectively as $\caseset$.

@@

In order to incorporate the determination of which case occurs into the
position and velocity update equations, we introduce an indicator function
$\indic_{t+1}^{\casegen}$ for each case $\casegen \in \caseset$.  When
$\casegen$ corresponds to the case actually taken by PSO,
$\indic_{t+1}^{\casegen}$ evaluates to 1; otherwise it evaluates to 0.  We can
then sum over all of the cases, and the indicator function will make all of the
terms drop to zero except for the case that actually occurs.  For example, the
indicator function for the specific case $\casexn$ can be written as follows:

@@

@@

We ran experiments to compare our speculative parallelization of PSO to the
na\"ive parallelization.  At each iteration of the algorithms, we use one
processor to perform one function evaluation for one particle, be it
speculative or non-speculative.  The speculative version actually performs
two iterations of PSO at each ``iteration,'' so we instead call each
``iteration'' a ``round of evaluations.''

>>>>

%% PSO Stuff
\DeclareMathOperator*{\argmin}{arg\;min}
\DeclareMathOperator*{\argmax}{arg\;max}
\DeclareMathOperator*{\arginf}{arg\;inf}
\DeclareMathOperator*{\argsup}{arg\;sup}
\providecommand{\pers}{\ensuremath{P}}
\providecommand{\neigh}{\ensuremath{N}}
\providecommand{\leftind}{\ensuremath{L}}
\providecommand{\rightind}{\ensuremath{R}}
\providecommand{\nURand}{\ensuremath{U^\neigh}}
\providecommand{\pURand}{\ensuremath{U^\pers}}
\providecommand{\ppos}{\ensuremath{\Vec{x}}}
\providecommand{\fppos}{\ensuremath{\Vec{fx}}}
\providecommand{\pvel}{\ensuremath{\Vec{v}}}
\providecommand{\nbest}{\ensuremath{\Vec{b}^\neigh}}
\providecommand{\fnbest}{\ensuremath{\Vec{fb}^\neigh}}
\providecommand{\pbest}{\ensuremath{\Vec{b}^\pers}}
\providecommand{\fpbest}{\ensuremath{\Vec{fb}^\pers}}
\providecommand{\constriction}{\ensuremath{\chi}}
\providecommand{\ncoeff}{\ensuremath{\phi^\neigh}}
\providecommand{\pcoeff}{\ensuremath{\phi^\pers}}
\providecommand{\obs}{\ensuremath{\Vec{\xi}}}
\providecommand{\ofunc}{\ensuremath{f}}
\providecommand{\swarm}{\ensuremath{swarm}}

@@

Particle swarm optimization (PSO) has been found to be a highly robust and
effective algorithm for solving many types of optimization
problems~\citep{poli-2008-pso-applications}.  For much of the algorithm's
history, PSO was run serially on a single machine.  However, the world's
computing power is increasingly coming from large clusters of processors.  In
order to efficiently utilize these resources for computationally intensive
problems, PSO needs to run in parallel.

@@

However, it can be seen from the graph that once the swarm is sufficiently
large, doubling the swarm size produces only incremental benefit; performance
improves, but not by very much.  The reason for this is that there comes a
point of diminishing returns with respect to adding particles.  In
\fig{evals-sphere} we show the value obtained after 50,000 function evaluations
(not iterations) as a function of swarm size, again for the function Sphere.
Increasing the swarm size from 5 to 10 has a significant effect on the value
obtained.  However, increasing the swarm size from 16 to 30 makes the algorithm
less efficient; that is, it reduces the progress the algorithm makes per
evaluation.  Other functions show similar trends, though often the optimal
swarm size is slightly larger.  For this reason, previous work has recommended
the use of a swarm size of 50 for
PSO~\citep{bratton-2007-defining-a-standard-for-pso}.  Thus, in at least some
cases, adding particles indefinitely will not yield an efficient
implementation.

@@

In this paper we consider PSO parallelization strategies for clusters of
hundreds or thousands of processors and functions for which a single evaluation
will take at least tens of seconds, but probably minutes or perhaps hours.  Our
purpose is to explore the question of what to do with a thousand processors
when 50 or 100 particles is the most efficient swarm size, and simply adding
particles results in only incremental improvement.  In our work we assume that
the researcher has some fixed number of processors that should be fully
utilized throughout the optimization process.

In order to solve this problem, we apply the concept of speculative
decomposition~\citep{grama-2003-intro-to-parallel-computing} to particle swarm
optimization, using extra processors to perform two iterations of PSO at the
same time.  Speculative decomposition is analogous to speculative execution
(also known as branch prediction), a technique commonly used in processors.
Modern processors, when faced with a branch on which they must wait (e.g., a
memory cache miss), guess which way the branch will go and start executing,
ensuring that any changes can be undone.  If the processor guesses right,
execution is much farther ahead than if it had idly waited on the memory
reference.  If it guesses wrong, execution restarts where it would have been
anyway.  Thus the processor speculates about future paths of execution in an
attempt to decrease overall processing time.

In this paper we show that the results of standard PSO can be reproduced
\emph{exactly}, two iterations at a time, using a speculative approach adapted
from speculative execution. We prove that the standard PSO equation can be
factored such that a set of speculative positions can be found which will
always include the position computed in the next iteration.  By computing the
value of the objective function for each of the speculative positions at the
same time the algorithm evaluates the objective function for the current
position, it is possible to know the objective function values for both the
current and the next iteration at the same time.  We demonstrate this principle
by implementation and show that it produces exactly the same results as
standard PSO, but two iterations at a time.  The resulting implementation runs
efficiently on large clusters where the number of processors is much larger
than a typical or reasonable number of particles, producing better results in
less ``wall-clock'' time.

It is important to note here that this is not a variant of PSO.  We simply
propose a new way to think about the parallelization of PSO that we show
performs better in many cases than previous parallelizations.

@@

The balance of this paper is organized as follows. \sec{pso} describes the
particle swarm optimization algorithm, and \sec{related} gives a brief overview
of previous parallelization techniques for this algorithm.  \sec{sepso}
describes mathematically how speculative evaluation can be done in parallel PSO
to perform two iterations at once, then \sec{implementation} gives information
on how to implement this method.  In \sec{relax}, we discuss various methods of
improving the performance of speculative evaluation in PSO, all of which break
the requirement of strictly reproducing the behavior of the original algorithm.
\sec{setup} describes the experiments we ran, and \sec{results} presents the
results of those experiments.  In \sec{conclusion} and \sec{future} we conclude
and discuss future work.

@@

At each iteration of constricted PSO, the position $\ppos_t$ and velocity
$\pvel_t$ of each particle are updated as follows:
\begin{align}
\nonumber
	\pvel_{t+1} &=
		\constriction \bigl[ \pvel_t
			+ \pcoeff\pURand_{t}\otimes(\pbest_{t} - \ppos_{t}) \\
\label{eq:velupdate}
			& \quad \quad \quad \, + \ncoeff\nURand_{t}\otimes(\nbest_{t} - \ppos_{t})
		\bigr] \\
\label{eq:posupdate}
	\ppos_{t+1} &= \ppos_{t} + \pvel_{t+1}
\end{align}
where \( \pURand_{t} \) and \( \nURand_{t} \) are vectors of independent random
numbers drawn from a standard uniform distribution, the \( \otimes \) operator
is an element-wise vector multiplication, $\pbest$ (called personal best) is
the best position the current particle has seen, and $\nbest$ (called
neighborhood best) is the best position the neighbors of the current particle
have seen~\citep{bratton-2007-defining-a-standard-for-pso}.  The parameters \(
\ncoeff \), \( \pcoeff \), and \( \constriction \) are given prescribed values
required to ensure convergence (2.05, 2.05, and .73,
respectively)~\citep{clerc-2002-constricted-pso}.

@@

The ideal topology and swarm size for PSO depend on the objective function.
Researchers have devised various benchmark functions and have found that the
ideal topology for one function may perform very poorly for another function.
The No Free Lunch Theorems for Optimization show that this is true in
general---if an algorithm performs well on average for one class of functions
then it must do poorly on average for other
problems~\citep{wolpert-1997-nfl-for-optimization}.

An attempt to standardize PSO found that although a Complete swarm of 50
particles converged to the global optimum more quickly for many benchmark
functions than the same swarm size with a Ring topology, it was also more
likely to permaturely converge to local optima for other functions.  The study
found no significant improvement for any other swarm size between 20 and 100
and concluded with a recommendation to use a swarm of 50 particles with a Ring
topology as a starting point.  The authors acknowledged that choosing the ideal
topology requires thorough experimentation for the particular
problem~\citep{bratton-2007-defining-a-standard-for-pso}.  Other authors have
explained that a large swarm with a sparse topology propogates information
slowly~\citep{montes-de-oca-2009-frankensteins-pso}.

\section{Related Work}
\label{sec:related}

There have been several parallel implementations of PSO presented in the
literature.  These implementations vary along two axes---what computation is
performed and how it is done.  How the computation is performed involves both
architectural decisions, such as a master-slave versus a fully distributed
architecture, and decisions about synchronicity.  What computation is performed
involves decisions about how to scale the PSO algorithm for parallel
implementations, such as the ratio of particles to processors or the topology
to use.

\subsection{Synchronous vs. Asynchronous PSO}

Synchronous parallel implementations of PSO reproduce the standard serial
algorithm exactly.  This approach was first described analytically by
\citet{belal-2004-parallel-models-for-pso} and first implemented by
\citet{schutte-2004-parallel-global-optimization-with-pso}.  In a typical
master-slave algorithm, the master assigns tasks to slave processors, and in
parallel PSO, each task consists primarily of a function evaluation.  Updating
the particle's position and value may also be included in the
task~\citep{belal-2004-parallel-models-for-pso}, or this work may be performed
in serial by the
master~\citep{schutte-2004-parallel-global-optimization-with-pso}.  Before
proceeding to the next iteration, particles communicate, and each particle
updates its neighborhood best.  Whether this communication step happens
sequentially on the master or in parallel, each particle must receive
communication from its neighbors before proceeding.  The benefits of the
synchronous PSO include its simplicity, repeatability, and comparability with
standard PSO, which may be essential in research applications.

Asynchronous parallel
PSO~\citep{venter-2005-parallel-pso-asynchronous-evaluations,
koh-2006-parallel-asynchronous-pso} is a modification to the standard algorithm
which removes the synchronization point at the end of each iteration.  Instead,
particles iterate independently and communicate asynchronously.  In a typical
master-slave implementation of asynchronous parallel PSO, the master updates
each particle's personal best, neighborhood best, velocity, and position
immediately after receiving the function value from the slave processor.  Since
this update occurs while other particles are still being evaluated, it may use
information from the previous iteration for some
neighbors.\footnote{Asynchronous parallel PSO has been compared to the
``asynchronous updates'' variant of serial
PSO~\citep{koh-2006-parallel-asynchronous-pso}.  However, serial PSO with
asynchronous updates differs from standard PSO in that particles use newer
information, but asynchronous parallel PSO differs from standard PSO in that
particles use older information.} In a partially asynchronous implementation,
particles might wait for some but not all neighbors to complete before
proceeding~%
\citep{scriven-2008-asynchronous-pso-in-unreliable-distributed-environments}.
In some master-slave implementations, particles never get more than one
iteration ahead of
others~\citep{venter-2005-parallel-pso-asynchronous-evaluations,
koh-2006-parallel-asynchronous-pso}.  However, in a fully distributed
implementation, particles might never wait for information, and one particle
could complete many more iterations than another
particle~\citep{scriven-2008-distributed-pso-using-peer-to-peer}.  The main
effect of asynchronous evaluation is that processors spend less time
idle---this trait is particularly valuable when processors are heterogeneous or
function evaluation times are
varied~\citep{venter-2005-parallel-pso-asynchronous-evaluations,
koh-2006-parallel-asynchronous-pso}.  Asynchronous parallel PSO behaves
differently than the standard algorithm and may even produce different results
between runs.  Most reports conclude that asynchronous communication produces
similar numerical results to the standard algorithm, but the question has not
yet been thoroughly
addressed~\citep{venter-2005-parallel-pso-asynchronous-evaluations,
koh-2006-parallel-asynchronous-pso}.

\subsection{Scaling PSO to many processors}

The contribution of this work deals not with the implementation details of
architecture and synchronicity, but with what should be done with the PSO
equations when many hundreds or thousands of processors are available.  The
main issues that have been addressed in this space are how many particles to
use for a particular number of processors and what communication topology
should be employed.

The number of particles per processor has typically been decided by how long it
takes to evaluate the function being optimized.  When the function takes longer
than a few seconds to evaluate, previous techniques have assigned the number of
particles in the swarm to be the number of processors available~%
\citep{jin-2005-pso-antenna-designs,mcnabb-2009-large-particle-swarms},
advocating using as many processors as possible to get the best performance.
When the function takes less time to evaluate than the TCP/IP stack used to
send interprocessor communication, parallel implementations assign several or
many particles to a single processor~\citep{chu-2006-intelligent-parallel-pso,
chang-2005-parallel-pso-with-communication-strategies}.  Often the processor
only sends information about the best particle it evaluated to other
processors~\citep{mcnabb-2009-large-particle-swarms}.

Another popular method is simply to run PSO independently on each of the
processors available, taking the best result when all of the runs complete.  It
should be noted that this is equivalent to the previously stated method of
assigning many particles to each processor, only with no communication between
processors instead of little communication.  Both of these methods can be
described as changes in the communication topology of the original PSO
algorithm~\citep{mcnabb-2009-large-particle-swarms}.

Thus previous work in parallelizing PSO has consisted entirely of increasing
the swarm size and adapting the topology to be better suited to parallel
computation.

@@

\subsection{Refactoring the PSO Equations}

To perform two iterations at a time we must first refactor PSO such that the
determination of the value of the objective function is separate from the rest
of the computation.  For simplicity, this discussion will describe the case
where PSO is performing function minimization using the Ring topology.  In this
example, each particle has two neighbors, the ``right neighbor'' and ``left
neighbor,'' whose positions are represented as $\rightn$ and $\leftn$
respectively.  Though we will only describe the case of the Ring topology, this
method can easily be extended to arbitrary topologies.

The refactoring hinges on the idea that once the random coefficients
$\pURand_{t}$ and $\nURand_{t}$ are drawn, there are only a few possible new
positions, or updates, for $\nbest$ and $\pbest$.  For the Ring topology there
are 7 possible update cases, identified in Table~\ref{tab:evals}.  We label
each case with an identifier referring to the source of the update: a minus
sign ($-$) represents no update, $L$ represents an update to $\nbest$ coming
from the left neighbor, $R$ represents an update to $\nbest$ coming from the
right neighbor, and $S$ represents an update to either $\pbest$ or $\nbest$
coming from the particle itself.  As an example, $\casexn$ refers to the case
that the particle finds a new personal best, but neither it nor its neighbors
find a position that updated its neighborhood best.  In the equations that
follow, we refer to an unspecified update case as $\casegen$, and to the set of
cases collectively as $\caseset$.

@@

In order to incorporate the determination of which case occurs into the
position and velocity update equations, we introduce an indicator function
$\indic_{t+1}^{\casegen}$ for each case $\casegen \in \caseset$.  When
$\casegen$ corresponds to the case actually taken by PSO,
$\indic_{t+1}^{\casegen}$ evaluates to 1; otherwise it evaluates to 0.  We can
then sum over all of the cases, and the indicator function will make all of the
terms drop to zero except for the case that actually occurs.  For example, the
indicator function for the specific case $\casexn$ (which, as is shown in
Table~\ref{tab:evals}, means that the particle's personal best was updated, but
its neighborhood best was not) can be written as follows:

@@

\subsection{Discussion}

@@

We ran experiments to compare our speculative parallelization of PSO to the
na\"ive parallelization.  At each iteration of the algorithms, we use one
processor to perform one function evaluation for one particle, be it
speculative or non-speculative.  The speculative version actually performs
two iterations of PSO at each ``iteration,'' so we instead call each
``iteration'' a ``time step.''
